\chapter{Minimizing Potential Energy}
\label{chap:minimizing-potential-energy}

Potential energy is the capacity for doing work which results only from a system's configuration, \ie{} its particles' positions. Recall that we can express all particle positions ${\vec{r}_i}$ as a function of the generalized coordinates; hence the potential energy ${U}$, too, is a function of the generalized coordinates. In the following, we shall assume the potential energy function to be continuous and locally differentiable.

In conservative systems, mechanical energy is conserved, which is equivalent to the work done by a force being independent of its trajectory. Therefore we can write the forces in terms of the potential energy as
%
\begin{equation}
  \vec{F}_i \coloneqq -\frac{\partial U}{\partial \vec{r}_i}.
  \label{eqn:force-as-differential}
\end{equation}
%
The forces ${\vec{F}_i}$ are defined such that the potential energy ${U}$ is equal to the work one must do \emph{against} the forces to transfer an arbitrary constant reference configuration into the current configuration; hence the negative sign. As a result, the forces are always directed towards a lower energy level and therefore act to reduce the system's potential energy \cite{Nave}.

\hfill

\noindent
A particle is said to be in \emph{mechanical equilibrium} if there is no net force on said particle. Similarly, a particle system is in mechanical equilibrium if the net force on all its particles is zero \cite{Orear}. Using \cref{eqn:force-as-differential}, we find that a system is in equilibrium if and only if the gradient ${\nabla U}$ is zero for its current configuration. For (local) minima of the potential energy, the system is said to be in \emph{stable equilibria}.

Recall that the forces ${\vec{F}_i}$ are defined to act to reduce the system's potential energy. If a system has been displaced from an equilibrium configuration, said forces consequently act to return the system to equilibrium and are therefore called \emph{restoring forces}.

Oftentimes a system has multiple stable equilibria in which algorithms looking to minimize the system's potential energy can get stuck. It is a challenging problem to find a configuration in which the system's energy reaches a global minimum.





\clearpage
\input{Sources/03-01-Force-Directed-Algorithms}

\clearpage
\input{Sources/03-02-Explicit-Energy-Function}
